Grounding the interaction: anchoring situated discourse in everyday human-robot interaction

### 2011-07-15 paper_knowledge.tex
Integrated Raquel changes on ORO

\paragraph{Contextualizing} The information contextualizing is the ability for
a cognitive system to connect a fact with a \emph{cultural context}, an
\emph{interpretive scope} and the set of other facts previously acquired by
the agent.
%~ Since machines are limited to syntactic (in contrast to semantic)
%processing, we are mostly looking for a syntactic (\ie, based on symbols)
%matching between concepts representations (in our case, sets of alphanumeric
%characters).\fxfatal{il faut sans doute évoquer ici la relation
%sémantique/syntactique que propose Choamsky}.

We call \textit{cultural context} a broad set of common, general facts that are
considered as widely accepted amongst the interactors (like \textit{bottles may
contain water}). This knowledge is often referred as \emph{common-sense
knowledge}.

By \emph{interpretive scope}, we mean that a concept may have different
interpretation depending on the agent, the current situation or the time frame
the statement belongs to. Since a fact in one scope can be different (or even
inconsistent) with a fact in another scope (for instance, one object can be
visible for the robot and invisible for another agent), the underlying
knowledge representation system must properly handle these interpretive
frameworks.

@@

\paragraph{Synthesis} The synthesis of knowledge is the identification process
of facts (and their components, concepts and predicates) with each others. For
instance, if the robot observes a human sitting down at a table, and at the
same time, we tell it that ``Peter is sitting at the table'', we would like the
robot to infer that ``Peter'' may be the name of the human. \textit{Synthesis}
refers to the fact that several, \textit{a priori} non correlated, facts must
be associated to the same common concept. This process requires the ability to
control the logical consistency of the knowledge corpus. For instance, if we
add to the previous example the fact that the human that is sitting is a woman,
the synthesis ``Peter is the name of the human'' is not valid anymore.

\paragraph{Domain of validity} The domain of validity of an information,
lastly, covers several aspects: temporal, situational and probabilistic. The
main role of the domain of validity is to specify the scope in which an
information is (believed to be) true. While related to the previously concept
of \emph{interpretive scopes}, the domain of validity addresses the
question whether a fact must be or not considered in a given context. This
limitation of the validity is usually not carried by the fact itself. In the
previous example, for instance, the robot observes a human sitting at a table.
This fact ``a human is sitting at the table'' is true only for a limited
period, until the human stand up. This period of time is not directly
accessible (the robot does not know how long the human plans to stay), but the
knowledge representation system must be able to deal with this uncertainty and
should explicitly label this fact as being limited in time.
%~ ~ To know if a fact is permanent or transitional (as defined by Pollock
%\fxfatal{citation de Langage et Cognition, page 50}) is difficult (especially
%considering that a feature may be considered as permanent or not depending of
%the context: within the situation ``a familly meal''\fxfatal{Check
%translation!}, the fact ``the human is sitting at the table'' could be
%considered as permanent. Conversely, ``ground is static'' is generally
%considered as a static fact, expect if we are talking of planetary mechanics
%for instance. The difficuly lies in the selection of the relevant situation in
%which reasonning must be carried out at a given time) and have currently to be
%defined in the cultural background of the robot.

Those three aspects translate into a set of features that are desirable for a knowledge representation system:
\begin{itemize}
	\item represent raw information,
	\item render a general cultural background, in form of common-sense knowledge,
	\item attach interpretive scopes to new statements,
	\item add and connect new statements to knowledge already present,
	\item store restrictions on the domain of validity of the knowledge.
\end{itemize}

Besides, active processes would be desirable to:
\begin{itemize}
	\item acquire and maintain knowledge perceived from the physical world or
	retrieved from other sources (interaction with other agents, web-based contents,...)
	\item synthesize facts as much as possible,
	\item monitor contexts and manage accordingly the validity of the stored knowledge,
	\item ensure the logical consistency of the knowledge repository, and
	explicit inconsistencies when required\footnote{One may argue that the real world is 
	inherently inconsistent. In this article, we however make a
	\textit{consistent world} assumption, in order to leverage reasoning
	capabilities of the first-order logics. This is supported by the natural
	tendency of humans themselves to provide a consistent explanation of their
	world.}.
\end{itemize}

This list does not cover all the possible features that could be exposed by a
symbolic knowledge management system. Bio-inspired memory management ---the
ability to forget or reinforce knowledge--- or curiosity ---the ability to
identify where knowledge is lacking and actively trigger behaviours to acquire
it---, to give some examples, could arguably be added to the list. However,
this first analysis sets a convenient reference frame to understand and
evaluate knowledge representation systems, including the \textsc{ORO} knowledge
management system we propose.

\begin{figure}[!ht]
\centering
  \includegraphics[width=\columnwidth]{images/snapshot_oroview.png}
  \caption{Live snapshot from the \emph{oro-view} viewer depicting the
  knowledge model of the Jido robot during an experiment. Dots represent
  instances (ABox) while circles pyramids are classes (TBox)}
  \label{fig|oro-view}
\end{figure}

@@

The ORO platform~\cite{Lemaignan2010} is primarily designed as a central
knowledge storage service implemented as a server: at run-time, the robot
components can add or query statements. Figure~\ref{fig|oro-overview}
illustrates the overall architecture. The \emph{front-end} accepts and manages
connections from client components. The clients' requests are processed by a
set of internal modules: basic operations on statements, but also higher
cognitive and human-robot interaction related functionalities are available
(detailed thereafter). External plugins can also be easily hooked. The modules
rely on several parallel ontology \emph{back-ends} where the knowledge is
actually stored.

@@

The ORO server can be loaded with an initial set of statements which we call
the \emph{OpenRobots Common Sense Ontology}. It defines a small set of concepts
(and implicitly, a vocabulary) that can be used by all the modules of the robot
to unambiguously add or query facts. Moreover, the same ontology declares rules
and logical properties that are later used for inference.

The \emph{OpenRobots Common Sense Ontology} defines a small set of classes (56 are
currently defined) and predicates (60 are currently defined) focused on
concepts useful for robot interaction with humans. It includes both very broad
categories like \concept{SpatialThing}, \concept{Event} or \concept{Action}, and much more
concrete concepts as \concept{Table}, \concept{Book} or colors. Predicates allow to
describe the state of the agents and the world with relations like \concept{isOn},
\concept{sees}, \concept{currentlyPerforms}, etc.

@@

Our common sense ontology is closely aligned on the open-source
\textsc{OpenCyc}\footnote{\url{http://www.opencyc.org}} upper ontology.
\textsc{OpenCyc} defines a large taxonomy of concepts and semantic
relationships between concepts that are used in several other projects
(\textsc{WordNet, DBpedia}). This potentially eases the exchange and addition
of knowledge from these other sources. Relying on the \textit{de facto}
\textsc{OpenCyc} standard enables as well knowledge exchange with other robots
(for instance, the works previously mentioned by Daoutis and Tenorth rely on
the same \textsc{Cyc} concepts).

@@

Besides storing and reasoning about knowledge, we have developed in ORO several
features to manage knowledge at higher level: independent cognitive models for
each agent the robot interacts with, categorization capabilities, different
profiles of memory and events registration.

\subsubsection{Base functionalities}
\label{base}
ORO offers an extended set of methods for processing of facts at the triples level. It includes:
\begin{itemize}
	\item statement ({\it i.e.}, RDF triples) insertion, removal, update,
	\item pattern-based removal of statements,
	\item consistency check, insertion of statements with consistency
	constraint (only if the new fact does not lead to inconsistencies),
	\item fast concept lookup, with multi-lingual support,
	\item queries with patterns (for instance, \stmt{ * isOn
	table}, which means ``return me all objects on table'') and filters (for
	instance, \concept{weight < 150.0}),
	\item standard SPARQL\footnote{\url{http://www.w3.org/TR/rdf-sparql-query/}} queries.
\end{itemize}

These basic functionalities enable the incremental construction (and
exploitation) of a consistent knowledge model, and the \emph{Common Sense Ontology}
core set of concepts helps to build assertions that are related to previous one by offering
a predefined vocabulary.

@@

As pictured in Figure~\ref{fig|oro-overview}, ORO stores independent cognitive
models for each agent it interacts with. When ORO actually identifies a new
agent (or infers that some instance is an agent), it automatically creates a
new, separate, in-memory OWL model. Robot's components like supervision or
situation assessment may then store the agents' beliefs in a
separate model. All knowledge processing functions available on the robot
primary model are equally available in every agent model. This allows to store
and reason on different (and possibly globally inconsistent) models of the
world.

This feature actually allows us to consider the robot to be endowed with a
\emph{theory of mind}~\cite{Scassellati2002}: the robot can explicitly model
the belief state of its interactors, opening new possibilities for the control
architecture. We present in the section~\ref{dialog} on natural language
grounding how we use this feature to make sense of the human sentences from the
human own point of view.

@@

We have implemented several algorithms (common ancestors, computation of the
best discriminant, see~\cite{Ros2010b}) to help the robot cluster a set of
concepts based on their symbolic similarities (common properties, common
ancestors). One particular application of these functions is discrimination:
when interacting with a user, the robot often needs to clarify an ambiguity
produced by its human partner. For instance, a user may refer to a ``bottle''
whereas two bottles are currently visible: the discrimination routines can
identify possible (symbolic) differences (maybe the color or the size of the
bottles) that permit the robot to ask an accurate question to the user to solve
the ambiguity. This discrimination can occur from the robot perspective or from
a specific agent perspective. Usage of these categorization abilities are
illustrated in the \emph{Spy Game} experiment (section~\ref{spygame}).

@@

\subsubsection{The events framework} \label{events} Lastly, ORO allows external
modules to be triggered when specific events occur. For instance, when a
logical sentence becomes true or false, or if a new instance of a certain class
is added. One immediate application is reactive supervision: a component could
for instance subscribe to events of kind \setstmt{?agent isVisible true, ?agent type Human}. As soon as the perception layer detects a human in the
robot's field of view and accordingly updates the knowledge base, the
supervision component would be triggered back.

The event framework also takes advantage of the inference capabilities of ORO.
Thus an event can be indirectly triggered if its triggering conditions can be
inferred to be true.

Next sections present how symbolic knowledge is actually produced and added to
ORO.

>>>>

\paragraph{Contextualizing} is the ability for a cognitive system to connect a
fact with a \emph{cultural context}, an \emph{interpretive scope} and the set
of other facts previously acquired by the agent.
%~ Since machines are limited to syntactic (in contrast to semantic)
%processing, we are mostly looking for a syntactic (\ie, based on symbols)
%matching between concepts representations (in our case, sets of alphanumeric
%characters).\fxfatal{il faut sans doute évoquer ici la relation
%sémantique/syntactique que propose Choamsky}.

We call \textit{cultural context} a broad set of common, general facts that are
considered widely accepted among the inter-actors (\eg ``bottles may contain
water''). This knowledge is often referred as \emph{common-sense knowledge}.

By \emph{interpretive scope} we mean that a concept may have different
interpretations depending on the agent, the current situation or the time frame
the statement belongs to. Since a fact in one scope can be different (or even
inconsistent) with a fact in another scope (for instance, one object can be
visible for the robot and invisible for another agent), the underlying
knowledge representation system must properly handle these interpretive
frameworks.

@@

\paragraph{Synthesis} corresponds to the identification of facts and their
components (concepts and predicates) with respect to other facts. For instance,
if the robot observes a human sitting down at a table, and at the same time, we
tell it that ``Peter is sitting at the table'', we would like the robot to
infer that ``Peter'' may be the name of the human. \textit{Synthesis}
refers to the fact that several, \textit{a priori} non correlated, facts
must be associated to the same common concept. This process requires the
ability to control the logical consistency of the knowledge corpus. To
continue with the previous example,  if we add the fact that the human that
is sitting is a woman, the synthesis ``Peter is the name of the human'' is
not valid anymore.

\paragraph{Domain of validity} its main role is to specify the scope in which
an information is (believed to be) true. It covers several aspects: temporal,
situational and probabilistic. While related to the previous concept of
\emph{interpretive scopes}, the domain of validity addresses the question
whether a fact must be or not considered in a given context. This validity
limitation is not usually carried by the fact itself. In the previous example,
for instance, the robot observes a human sitting at a table.  The fact ``a
human is sitting at the table'' is true only for a limited period of time,
until the human stands up. This period of time is not directly accessible
(the robot does not know how long the human plans to stay), but the
knowledge representation must be able to deal with this uncertainty and
should explicitly label this fact as being limited in time.
%~ ~ To know if a fact is permanent or transitional (as defined by Pollock
%\fxfatal{citation de Langage et Cognition, page 50}) is difficult (especially
%considering that a feature may be considered as permanent or not depending of
%the context: within the situation ``a familly meal''\fxfatal{Check
%translation!}, the fact ``the human is sitting at the table'' could be
%considered as permanent. Conversely, ``ground is static'' is generally
%considered as a static fact, expect if we are talking of planetary mechanics
%for instance. The difficulty lies in the selection of the relevant situation
%in which reasoning must be carried out at a given time) and have currently to
%be defined in the cultural background of the robot.

These three aspects lead us to envisage a knowledge representation system
characterized by the following features: 
\begin{itemize}
	\item represent raw information,
	\item render a general cultural background, in the form of common-sense knowledge,
	\item attach interpretive scopes to new statements,
	\item add and connect new statements to knowledge already present,
	\item store restrictions on the domain of validity of the knowledge.
\end{itemize}

Besides, the following active processes would be desirable to:
\begin{itemize}
	\item acquire and maintain knowledge perceived from the physical world or
	retrieved from other sources (interaction with other agents, web-based contents,...)
	\item synthesize facts as much as possible,
	\item monitor contexts and accordingly manage the validity of the stored knowledge,
	\item ensure the logical consistency of the knowledge repository, and
	explicit inconsistencies when required\footnote{One may argue that the real world is 
	inherently inconsistent. In this article, we however make a
	\textit{consistent world} assumption, in order to leverage reasoning
	capabilities of the first-order logics. This is supported by the natural
	tendency of humans themselves to provide a consistent explanation of their
	world.}.
\end{itemize}

This list does not cover all the possible features that could be exposed by a
symbolic knowledge management system. Bio-inspired memory management ---the
ability to forget or reinforce knowledge--- or curiosity ---the ability to
identify lacking knowledge and actively trigger behaviors to acquire
it\fixme{cite Howa}---, to give some examples, could arguably be added to the
list. However, this first analysis sets a convenient reference frame to
understand and evaluate knowledge representation systems, including the
\textsc{ORO} knowledge management system we propose.

\begin{figure}[!ht]
\centering
  \includegraphics[width=\columnwidth]{images/snapshot_oroview.png}
  \caption{Live snapshot from the \emph{oro-view} viewer depicting the
  knowledge model of the robot during an experiment. Dots represent
  instances (ABox), while circles pyramids are classes (TBox)}
  \label{fig|oro-view}
\end{figure}

@@

The ORO platform~\cite{Lemaignan2010} is primarily designed as a central
knowledge storage service implemented as a server where the robot
components can add or query statements at run-time. Figure~\ref{fig|oro-overview}
illustrates the overall architecture. The \emph{front-end} accepts and manages
connections from client components. The clients' requests are processed by a
set of internal modules: basic operations on statements, but also higher
cognitive and human-robot interaction related functionalities are available
(detailed thereafter). External plugins can also be easily hooked. The modules
rely on several parallel ontology \emph{back-ends} where the knowledge is
actually stored.

@@

The ORO server can be loaded with an initial set of statements which we call
the \emph{OpenRobots Common Sense Ontology}. It defines a small set of concepts
(and implicitly, a vocabulary) that can be used by all the modules of the robot
to unambiguously add or query facts. Moreover, the same ontology declares rules
and logical properties that are later on used for inference.

The \emph{OpenRobots Common Sense Ontology} defines a small set of classes (56
are currently defined) and predicates (60 are currently defined) focused on
concepts useful for human-robot interaction. It includes both very broad
categories like \concept{SpatialThing}, \concept{Event} or \concept{Action},
and much more concrete concepts as \concept{Table}, \concept{Book} or colors.
Available predicates allow to describe the state of the agents and the world
with relations like \concept{isOn}, \concept{sees},
\concept{currentlyPerforms}, etc.

@@

Our common sense ontology is closely aligned with the open-source
\textsc{OpenCyc}\footnote{\url{http://www.opencyc.org}} upper ontology.
\textsc{OpenCyc} defines a large taxonomy of concepts and semantic
relationships between concepts that are used in several other projects
(\textsc{WordNet, DBpedia}). This potentially eases the exchange and addition
of knowledge from these other sources. Moreover, it also enables knowledge
exchange with other robots (for instance, the works previously mentioned by
Daoutis and Tenorth rely on the same \textsc{Cyc} concepts\fixme{Give
examples}).

@@

Besides storing and reasoning about knowledge, we have developed in ORO several
features to manage knowledge at higher level: 
%independent cognitive models for
%each agent the robot interacts with, categorization capabilities, different
%profiles of memory and events registration.

\subsubsection{Base functionalities}
\label{base}

ORO offers an extended set of methods to process facts at the triples level,
including:
\begin{itemize}
	\item statement (\ie RDF triples) insertion, removal, update,
	\item pattern-based statements removal,
	\item pattern-based queries (for instance, \stmt{ * isOn
	table}, which means ``return me all objects on table'') and filters (for
	instance, \concept{weight < 150.0}),
	\item consistency check, insertion of statements with consistency
	constraint (only if the new fact does not lead to inconsistencies),
	\item fast concept lookup, with multi-lingual support,
	\item standard SPARQL\footnote{{\em SPARQL Protocol and RDF Query Language}, \url{http://www.w3.org/TR/rdf-sparql-query/}} queries.
\end{itemize}

While these basic functionalities enable the incremental construction (and
exploitation) of a consistent knowledge model, the \emph{Common Sense Ontology}
helps building assertions that are related to previous one by offering
a predefined vocabulary.

@@

As pictured in Figure~\ref{fig|oro-overview}, ORO stores independent cognitive
models for each agent it interacts with. When ORO actually identifies a new
agent (or infers that some instance is an agent), it automatically creates a
new, separate, in-memory OWL model for that agent. Thus, different robot's
components, like supervision or situation assessment, may then store the
agents' beliefs in separate models. All knowledge processing functions in the
robot's primary model are equally available in every agent's model, which
allows to store and reason on different (and possibly globally inconsistent)
models of the world.

This feature actually allows us to consider the robot to be endowed with a
\emph{theory of mind}~\cite{Scassellati2002}: the robot can explicitly model
the belief state of its inter-actors, opening new possibilities for the control
architecture. In the section~\ref{dialog} we present how we use this feature to
make sense of the user sentences from the his/her point of view. Moreover,
these multiple models can be viewed as well as different interpretive scopes,
allowing to interpret the same reality from different points of view.

@@

We have implemented several algorithms (common ancestors, computation of the
best discriminant, see~\cite{Ros2010b}) to help the robot cluster a set of
concepts based on their symbolic similarities (common properties, common
ancestors). One particular application of these functions is discrimination.
While interacting with a user, the robot quite often needs to clarify an
ambiguity produced by its human partner. For instance, a user may refer to a
``bottle'' whereas two bottles are currently visible. Discrimination routines
can identify possible (symbolic) differences (\eg the color or the size of the
bottles) that permit the robot to ask an accurate question to the user in order
to solve the ambiguity. This discrimination can occur from the robot's
perspective or from a specific agent's perspective. Usage of these
categorization abilities are illustrated in the \emph{Spy Game}
(section~\ref{spygame}).

@@

\subsubsection{The events framework} 
\label{events}

Lastly, ORO allows external modules to be triggered when specific events occur.
For instance, when a logical sentence becomes true or false, or if a new
instance of a certain class is added. One immediate application is reactive
supervision: a component could for instance subscribe to events of kind
\setstmt{?agent isVisible true, ?agent type Human}. As soon as the perception
layer detects a human in the robot's field of view and accordingly updates the
knowledge base, the supervision component would be triggered back. The event
framework also takes advantage of the inference capabilities of ORO. Thus an
event can be indirectly triggered if its triggering conditions can be inferred
to be true.

Next sections describe how symbolic knowledge is actually produced and added to
ORO.
